---
title: "population_clarification"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, packages}
library(tidyverse)
library(scales)
library(gridExtra)
library(data.table)
library(ggmap)
library(randomForest)
library(gridExtra)
library(GGally)
library(sm)
source("https://raw.githubusercontent.com/janhove/janhove.github.io/master/RCode/sortLvls.R")
library(RColorBrewer)
library(ash)
library(lattice)
colors_vec <- c("forestgreen", "orange", "red")
```

```{r, include = F,message=FALSE, warning=FALSE, read_data}
train_x = read_csv('data/train_x.csv')
train_y = read_csv('data/train_y.csv')
test_x = read_csv('data/test_x.csv')
water = merge(train_x, train_y)
```

Below is a plot having proportions of functional status of waterpoints on the y-axis and   population (log transformed) on the x-axis. The population variable here is log 10 transformed so 1 on the x-axis corresponds to population 10, 2 corresponds to 100 people, 3 corresponds to 1000 people and 4 corresponds to 10,000 people and 4.48 corresponds to 30500 people which is the max. population in the data. 

Now in case of larger populations (which are around 1000 or greather than 1000) i.e. around 3 or greater than 3 on this plot you can see that the proportion of pumps that are `functional needs repair` as indicated by the orange vertical bar is larger than that of lower population pumps indicated by their smaller orange vertical bars.

Another thing we notice is that compared to very small population pumps (look at the red in the second vertical bar which represents the proportion of `non-functional` pumps), higher population pumps appear to be somewhat less likely to be `non functional` i.e more smaller population pumps are `non-functional` compared to higher population pumps.

Finally, contrary to the final point in the report, it appears that 10-100 people around a water point is most favorable in terms of functionality. That is we see that the proportion of `functional` pumps is highest in areas with 10-100 population as indicated by the 3rd, 4th, 5th, 6th, 7th and 8th bars.

```{r, echo =F,message=FALSE, warning=FALSE}
#normalize the counts per value level
normalizedCounts <- function(x, y, breaks = NULL, labels = NULL){
  if (class(x) == 'numeric' || class(x) == 'integer'){
    x <- cut(x, 
             breaks = breaks, 
             labels = labels)
  }
  df <- data.frame(x = x, y=y)
  tab <- as.data.frame(table(df))
  tab_wide <- reshape(tab, 
                      v.names="Freq", 
                      timevar="y", 
                      idvar="x", 
                      direction='wide')
  rownames(tab_wide) <- tab_wide$x
  tab_wide$x <- NULL
  tab_norm <- as.data.frame(t(apply(tab_wide, 1, function(x) x/sum(x))))
  if (class(x) == 'numeric' || class(x) == 'integer'){
    tab_norm$label <- as.numeric(rownames(tab_norm))
  }else{
    tab_norm$label <- rownames(tab_norm)
  }
  melt(tab_norm)
}

bivariate <- function(x, y, 
                      breaks = NULL, 
                      labels = NULL, 
                      facet = factor(), 
                      colors = colors_vec, 
                      names_vec = NULL) {
  
  df <- data.frame(x=x, y=y)
  rel_count_functional <- nrow(subset(df, y=='functional'))/nrow(df)
  
p2 <- ggplot(normalizedCounts(df$x, 
                                df$y,
                                breaks,
                                labels), 
               aes(x=label, y=value, fill=variable)) 
  

p2 <- p2 + geom_bar(stat='identity', aes(x=as.numeric(label))) + scale_x_continuous()
  
 
  p2 <- p2 + geom_hline(yintercept=rel_count_functional) +
             theme(axis.text.x = element_text(angle = 0, 
                                              hjust = 1, 
                                              vjust = 0.3), legend.position="right") +  scale_fill_manual(values= c("forestgreen", "orange", "red")) 
return(p2)  
}
  
  
            
```

```{r, echo = F, message=FALSE, warning=FALSE}
bivariate(log10(water$population + 1) , water$status_group,  breaks = seq(-0.25, 4.7, by = 0.25), labels = seq(-0, 4.7, by = 0.25))
```

